Making multiple-locus association mapping on a genome-wide scale routine

### 2016-11-01 AM+draft.tex
2/11/2017 amout to begin writing the results section

Over the past decade, g-w association mapping methods have changed considerably. Early mapping methods relied upon 
simple contingency tables or logistic regression for case-control designs and score tests for family-based designs.

Over the past decade, genome-wide association mapping methods, have changed considerably. Early association mapping 
methods were no more complicated than contingency table tests or logistic regression. This is because early association 
studies followed case-control designs. A significant difference in the allele frequency at a locus, between the cases (affecteds) and 
controls (unaffecteds), was interpreted as the locus being in linkage disequilibrium with a casual gene. 
It became known though that differences in allele frequency could also arise due to unrecognised population stratification. 
This prompted a shift towards family-based designs. Family-based designs do not suffer from 
the confounding affects of population stratification. A myriad of new family-based association mapping methods were developed, 
the most popular based on score tests (ref for TDT S-TDT Spielman and Ewens 1998, AJHG 62:450-458, FBAT Naird, NM; Horvath, S; Xu, X 2000, genetic epi V19, Suppl 1 S36-S42).

Over the past decade, genome-wide association mapping, the statistical techniques for analysing data from association studies, has advanced considerably. Early genome-wide association studies (GWAS) began as case-control designs. The premise was that casual genes could be detected via differences in allele frequency between cases (affecteds) and controls (unaffecteds). Association mapping methods for these designs were no more sophisticated than contingency table tests or logistic regression. Unfortunately, differences 
in allele frequency can also arise due to unrecognised population stratification. Consequently, there was a shift towards family-based 
designs. Findings from these designs are
not susceptible to the confounding affects of population stratification. With interest growing in genome-wide association studies, 
a myriad of different association mapping methods for family-based designs were developed.

Over the past decade, there have also been impressive advancements in genome wide 
association mapping methods, the statistical techniques for measuring the strength of 
association between a marker locus and trait. (Needs more detail - what are some of the early assoc study methods (give a single sentence review).... TDT, genomic control, structured association, score tests???? 
Traditionally, association data were 
analysed with simple contingency tests or Cochran-armitage  trend tests. Today, far 
more statistically sophisticated methods are available. One such method is the  QK 
approach, a linear mixed model framework for genome wide association 
mapping (refs).  The QK approach

Several highly efficient computer programs for performing QK-based 
association mapping are available. Programs such as EMMA, fastLMM, GEMMA, and 
XXX(ref) make feasible the analysis of data from large GWAS. 
 However, not all areas of genome-wide association 
mapping have advanced. It still remains standard practice to analyse genome-wide 
association data with a single locus model. That is,  it is common for the strength of 
association between a marker locus and trait is measured for each locus separately.

Methods for association mapping with multiple locus models have been developed 
for some time. Computer implementations of these methods have also been available. 
Thus, it is not a lack of methodology  or computational tools preventing the uptake of 
multiple locus association mapping. The reasons for their lack of acceptance are many. 
Multiple locus association mapping methods are often cased in high level statistical theory, 
making their true value difficult to appreciate.  Computationally, they can scale poorly with 
sample size and/or number of genetic markers. Some methods require tuning parameters 
to be specified. These methods also report their results as coefficients whose significance 
are unknown. There are multiple locus methods that instead report their results as probabilities 
but the difficulty here is determining  an appropriate threshold for inferring a statistically 
significant finding. The computer implementations can also be difficult to run in some cases. 
All these reasons contribute to making multiple locus association mapping unappealing as 
an analysis method.

In this paper, we present our new am+ approach for multiple locus association mapping. 
Our approach is based on linear mixed models and uses the same parameter estimation 
procedure as EMMA. Through a clever dimension reduction step, our approach scales 
with number of subjects instead of number of marker loci. We build the multiple locus 
model iteratively. At each step, only a single linear mixed model is fitted to the data. 
This model fits all marker loci and their association with the trait simultaneously. There 
are no parameters to be tuned, no thresholds to be set. Our approach automatically 
finds the best model. Once our approach is run, the set of statistically significant 
marker loci in strongest association with the trait is returned as its findings.

@@

@@

>>>>

Over the past decade, Genome wide association studies have changed considerably in both their analysis and design. Early gwas followed a case-control design. Association mapping methods were no more complicated than contingency table tests or regression. These designs though had a tendency to yield spurious findings if there was unrecognised population stratification. This prompted a shift towards family-based designs and score tests, such as the tdt test and its variants (refs), became popular. Today, instead of by design, it is through statistical modelling that we account for the effects of population stratification. This has meant that data can be collected from general populations, even if these populations are highly structured. Analysis via sophisticated association mapping methods based on linear mixed models is now almost routine.

What has not changed over the past decade is that it remains common practice to analyse gwas data on a locus by locus basis. This is despite there being several significant problems with analysing the data in this way. First, the aim of association mapping is to identify regions of the genome that house genes that are influencing a trait. The identification of these regions from these analyses is not always straightforward. Gwas results are reported, typically, via Manhattan plots that plot the -log10 of the p value for each locus against the map position of the locus. The location of peaks in this plot signify regions of interest. Inferring the exact number of regions of interest can be difficult If the peaks are not well separated. Second, when multiple statistical tests are performed, the probability of wrongly accepting a significant result (type 1 error) is inflated. This is known as the multiple testing problem (refs). Many different solutions have been offered (refs). Yet, there is still no well accepted way of correcting for multiple testing in the context of genome- wide association mapping. Third, many of the traits whose genetic secretes we are trying to discover are complex. There will be multiple loci in linkage equilibrium with genes that influence the trait. Yet, A locus by locus mapping approach only assesses the evidence for association between a single locus and trait.

It is somewhat surprising then that multiple locus methods haven't attracted more attention. It is possible to measure all locus-trait associations jointly with methods based on the regularisation techniques, ridge regression and lasso. Here, multiple testing is not an issue. These techniques though are computationally demanding. Also, the results are not easy to interpret. The strength of association is not measured by a p-value but by the size of the regression coefficient for a locus in the model. More recently, associations have started to be mapped with random forests (refs). Similar to regularisation techniques though, it is not clear how to infer genomic regions of interest from their findings (refs). An multiple locus method that does show promise is the multi-locus linear mixed model method (ref). The best multiple locus model is built with simple variable selection. Results are immediately interpretable but here, computation can be a challenge for large data.

Here, we present our new multiple locus method for genome wide association mapping, which we are calling AMplus. AMplus combines the strength of regularisation techniques (being able to fit all locus-trait associations jointly), with simple variable selection (having easy to interpret results). Our method does not require a significance threshold to be set nor regularisation parameters to be fine tuned. Through a clever dimension reduction step, we are able to achieve a computational performance similar to the fastest single locus linear mixed model implementations. AMplus is statistically more powerful than single locus association mapping and is as and often more powerful than 
most multiple locus methods. Our aim is to make multiple locus association mapping on a genome wide scale routine. To this end, we have created an R package ...

@@

\subsection{Time and Memory Performance}

@@

\subsection{Run times}

\subsection{Memory}
